---
title: "SURVIVAL ANALYSIS"
subtitle: "Applied Biostatistics Course with R"
author: "Biostatistics Support and Research Unit, IGTP"
format: 
  html:
    embed-resources: true
    toc: true
    code-overflow: wrap
execute: 
  echo: true
  eval: false
editor_options: 
  chunk_output_type: console
---

## Example dataset: Advanced lung cancer

```{r eval=TRUE, echo=FALSE, message=TRUE, warning=TRUE}
#install.packages("survival")
library(survival)
```

```{r}
#lung
head(lung,3)
```

## Example dataset

```{r}
library(dplyr)
lung <- 
  lung |> 
  mutate(
    status = case_when(
      status == 1 ~ 0,
      status == 2 ~ 1,
      .default = status
    )
  )
```


## Survival function

```{r}
surv_obj <- Surv(time = lung$time, event = lung$status)  # creates a survival object
head(surv_obj, 10)   # prints first 10 values   
```


## Kaplan-Meier method

```{r eval=TRUE}
km <- survfit(Surv(time = time, event = status) ~ 1, data = lung)  
```

```{r eval=TRUE}
#install.packages("survminer")
library(survminer)
ggsurvplot(km) #?ggsurvplot
```

```{r eval=TRUE}
ggsurvplot(km, xlab = "Days", legend = "none", censor=FALSE, conf.int=FALSE, risk.table=TRUE)
```

```{r eval=TRUE}
summary(km, time = c(365, 730))     # survival estimates at 1 and 2 years
```

```{r eval=TRUE}
km
```

```{r eval=TRUE}
quantile(km)
```

# Comparing survival curves

## Log-rank test in R

```{r eval=TRUE}
library(dplyr)
# define factor for sex 
lung <- lung |>
   mutate(sex = factor(sex, levels=1:2, labels = c("male", "female")))
# specify formula
survdiff(Surv(time, status) ~ sex, data = lung)                            
```

## Comparing survival curves in R

```{r eval=TRUE}
ggsurvplot(survfit(Surv(time, status) ~ sex, data = lung),xlab = "Days",censor=FALSE, conf.int=FALSE, risk.table=TRUE, pval=TRUE, pval.method=TRUE)
```

# Cox regression models

## Cox regression models in R (qualitative variable)

```{r eval=TRUE}
m1 <- coxph(Surv(time, status) ~ sex, data = lung)
m1
```

```{r eval=TRUE}
library(gtsummary)
coxph(Surv(time, status) ~ sex, data = lung) |> 
  tbl_regression(exp = TRUE)
```

## Cox regression models in R (quantitative variable)

```{r eval=TRUE}
coxph(Surv(time, status) ~ age, data = lung) |> 
  tbl_regression(exp = TRUE)
```

## Cox regression models in R (more than one variable)

```{r eval=TRUE}
coxph(Surv(time, status) ~ age + sex + ph.ecog, data = lung) |> 
  tbl_regression(exp = TRUE)
```

## Model validation

## Model validation: Proportional hazards test

```{r eval=TRUE}
m2 <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data = lung)
test.ph <- cox.zph(m2)
test.ph
```

```{r eval=TRUE}
m2 <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data = lung)
test.ph <- cox.zph(m2)
test.ph
```

## Model validation: Graphs of Schoenfeld residual

```{r}
ggcoxzph(test.ph)
```

## Model validation: Graphs of Schoenfeld residual

```{r eval=TRUE}
ggcoxzph(test.ph)[1]
ggcoxzph(test.ph)[2]
ggcoxzph(test.ph)[3]
```

## Model validation: Log-log plots

```{r eval=TRUE}
km2 <- survfit(Surv(time, status) ~ sex, data = lung)
ggsurvplot(km2, fun = "cloglog")
```

## Non-proportional hazards

```{r eval=TRUE}
m3 <- coxph(Surv(time, status) ~ sex + meal.cal, data = lung)
test.ph <- cox.zph(m3)
test.ph
```

```{r eval=TRUE}
ggcoxzph(test.ph)[2]
```
