---
title: "ANALYSIS OF VARIANCE"
subtitle: "Applied Biostatistics Course with R"
author: "Biostatistics Support and Research Unit, IGTP"
format: 
  html:
    embed-resources: true
    toc: true
    code-overflow: wrap
execute: 
  echo: true
editor_options: 
  chunk_output_type: console
---

# Hands-on ONE-WAY ANOVA

## Guinea Pigs: data

The Effect of Vitamin C dose on Tooth Growth in Guinea Pigs

```{r , warning=F,echo=TRUE, eval=FALSE}

library(dplyr) # Data managment
library(gtsummary) # Summary of results
library(ggplot2) # Graphics

data(ToothGrowth) # Data

ToothGrowth <- ToothGrowth |>
    mutate(dose=factor(dose))  # dose as factor

ToothGrowth |> 
    dplyr::select(-supp) |>
    gtsummary::tbl_summary(
        by=dose,
        type = all_continuous() ~ "continuous2", 
        statistic=all_continuous()~c("{mean} ({sd})",
        "{median} ({p25}, {p75})"),
        digits = all_continuous() ~ 2,
        label = list(len = "Length (cm)"))  
            
# Plot
ggplot(ToothGrowth, aes(x = dose, y =len,color=dose)) +
  geom_point(size = 4)+
  geom_boxplot()+
  labs(x = "Dose", y = "Length (cm)") +
  theme_minimal() +
  theme(legend.position = "none")

```


## Guinea Pigs: normality

```{r , warning=F,echo=TRUE, eval=FALSE}

#QQplot
ggplot(ToothGrowth, aes(sample = len)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "Q-Q Plots for Each Group", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  facet_wrap(~dose) +  
  theme_minimal() 

#Shapiro-Wilk
ToothGrowth |>
  group_by(dose) |>
  summarise(p_value = shapiro.test(len)$p.value)

```



## Guinea Pigs: homocedasticity


```{r , warning=F,echo=TRUE, eval=FALSE}

# Plot
ggplot(ToothGrowth, aes(x =dose, y = len, fill = dose)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Violin+Jittered: Variance Visualization",
       x = "Dose",
       y = "Length (cm)") +
  theme(legend.position = "none")

# Levene Test
library(car)  
leveneTest(len ~ dose, data = ToothGrowth)

# Bartlett Test
bartlett.test(len ~ dose, data = ToothGrowth)

```


## Guinea Pigs: anova


```{r , warning=F,echo=TRUE, eval=FALSE}

# ANOVA: dependent ~ independent
model_anova <- aov(len ~ dose, data = ToothGrowth)  

# Arrange results
results_anova <- broom::tidy(model_anova)
results_anova$term<-c("Between-groups","Within-groups")
names(results_anova)<-c("Terms","df","SS","MSQ","F","P-value")

# Tabulate results
results_anova |>
  gt::gt() |>
  gt::fmt_number(decimals = 3) |>
  gt::tab_header(title = "Guinea Pigs: 1-way Anova") 

```



## Guinea Pigs: post-hoc

```{r , warning=F,echo=TRUE, eval=FALSE}

# Tukey's Honest Significant Difference Test
TukeyHSD(model_anova)

# Bonferroni
pairwise.t.test(ToothGrowth$len, ToothGrowth$dose, p.adjust.method = "bonferroni")


```


```{r , warning=F,echo=TRUE, eval=FALSE}

# Dunnett's Test
library(multcomp)
summary(glht(model_anova, linfct = mcp(dose = "Dunnett"))) 

```



## Guinea Pigs: marginal means

```{r , warning=F,echo=TRUE, eval=FALSE}

library(emmeans) 

# Estimated marginal means
emm <- emmeans(model_anova, ~ dose)
print(emm)

# Pairwise comparisons with Tukey adjustment
pairs(emm, adjust = "tukey")



```


## Guinea Pigs: marginal means


```{r , warning=F,echo=TRUE, eval=FALSE}

# Plot
emmip(emm, ~ dose) +  
  theme_minimal() + 
  labs(title = "Estimated Marginal Means with CI", 
       y = "Estimated Mean Length", 
       x = "Dose")


# Plot estimated marginal means with confidence intervals
ggplot(as.data.frame(emm),  aes(x = dose, y = emmean, ymin = lower.CL, ymax = upper.CL)) +
  # Mean points
  geom_point(size = 4, color = "blue") +  
  # Confidence intervals
  geom_errorbar(width = 0.2, color = "blue") +  
  labs(title = "Marginal Means with 95% CIs",
      x = "Dose",
      y = "Estimated Tooth Length") +
  theme_minimal() 

```


## Guinea Pigs: residuals

```{r , warning=F,echo=TRUE, eval=FALSE}

# Extract Standardized residuals and fitted values
ToothGrowth$resid <- rstandard(model_anova)  
ToothGrowth$fitted <- fitted(model_anova)

# Normality
ggplot(ToothGrowth, aes(sample = resid)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "Q-Q Plot of Residuals", 
       x = "Theoretical Quantiles", 
       y = "Sample Quantiles") +
  theme_minimal() 

# Residuals vs Fitted plot to check variance patterns
ggplot(ToothGrowth, aes(x = fitted, y = resid))+
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal() 

```


## Guinea Pigs: Heterogeneity


**Welch’s correction**

```{r , warning=F,echo=TRUE, eval=T}

library(WRS2)
welch<-t1way(len ~ dose, data = ToothGrowth)  
print(welch)

```

**White’s correction**

```{r , warning=F,echo=TRUE, eval=T}

library(rstatix)
white<-anova_test(len ~ dose, data = ToothGrowth, white.adjust = TRUE)
print(white)

```


# Hands-on TWO-WAY ANOVA

## Guinea Pigs: data reloaded

The Effect of Vitamin C dose and **Supplement type** on Tooth Growth in Guinea Pigs


```{r , warning=F,echo=TRUE, eval=FALSE}

library(dplyr) # Data managment
library(gtsummary) # Summary of results
library(ggplot2)# Graphics

data(ToothGrowth) # data

ToothGrowth <- ToothGrowth |>
    mutate(dose=factor(dose),  # dose as factor
           supp=factor(supp))  # supp as factor

ToothGrowth |>
  tbl_strata(
    strata = supp,  .tbl_fun =  ~ .x |>
        tbl_summary(by = dose, 
        missing = "no",
        type = all_continuous() ~ "continuous2", 
        statistic=all_continuous()~c("{mean} ({sd})",
        "{median} ({p25}, {p75})"),
        digits = all_continuous() ~ 2,
        label = list(len = "Length (cm)"))|>
    .header = "**{strata}**, N = {n}"
  )
            


```


## Guinea Pigs: data reloaded

The Effect of Vitamin C dose and **Supplement type** on Tooth Growth in Guinea Pigs

```{r , warning=F,echo=TRUE, eval=FALSE}

            
# Plot
ggplot(ToothGrowth, aes(x = dose, y = len, color = supp)) +
  geom_point(size = 4, alpha = 0.3) +
  geom_boxplot(aes(x = dose, group = interaction(dose, supp)), width = 0.4, alpha = 0.5) +
  facet_wrap(~supp) +  # Separate plots for OJ and VC
  labs(x = "Dose", y = "Length (cm)") +  
  theme_minimal() 


```


## Guinea Pigs: normality


```{r , warning=F,echo=TRUE, eval=FALSE}

#QQplot
ggplot(ToothGrowth, aes(sample = len)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~dose*supp) +  
  labs(title = "Q-Q Plots for Each Group", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal() +
  
```


## Guinea Pigs: homocedasticity


```{r , warning=F,echo=TRUE, eval=FALSE}

# Plot
ggplot(ToothGrowth, aes(x =dose, y = len, fill = dose)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  theme_minimal() +
  facet_wrap(~supp) +
  labs(title = "Violin+Jittered: Variance Visualization",
       x = "Dose",
       y = "Length (cm)") +
  theme(legend.position = "none")

# Levene Test
library(car)  
leveneTest(len ~ dose*supp, data = ToothGrowth)

# Bartlett Test
bartlett.test(len ~ interaction(dose, supp), data = ToothGrowth)

```


## Guinea Pigs: anova


```{r , warning=F,echo=TRUE, eval=FALSE}

# ANOVA: dependent ~ independent
model_anova_2w <- aov(len ~ dose*supp, data = ToothGrowth)  

# Arrange results
results_anova_2w <- broom::tidy(model_anova_2w)
results_anova_2w$term<-c("Between-groups: Dose","Between-groups: Supp","Interaction","Within-groups")
names(results_anova_2w)<-c("Terms","df","SS","MSQ","F","P-value")

# Tabulate results
results_anova_2w |>
  gt::gt() |>
  gt::fmt_number(decimals = 3) |>
  gt::tab_header(title = "Guinea Pigs: 2-way Anova") 

```


## Guinea Pigs: marginal means


```{r , warning=F,echo=TRUE, eval=FALSE}

# Estimated marginal means
emm <- emmeans(model_anova_2w, ~ dose*supp)
print(emm)

pairs(emm, adjust = "tukey")

```


## Guinea Pigs: marginal means


```{r , warning=F,echo=TRUE, eval=FALSE}

# Plot estimated marginal means with confidence intervals

ggplot(as.data.frame(emm), aes(x = dose, y = emmean, color = supp, group = supp)) +
  geom_point(size = 4) +
  geom_line() +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.2) +
  scale_x_discrete(labels = c("0.5", "1", "2")) +
  labs(x = "Dose", y = "Estimated Marginal Mean", color = "Supplement") 
  theme_minimal() 


```

## Guinea Pigs: residuals


```{r , warning=F,echo=TRUE, eval=FALSE}

# Extract residuals and fitted values

ToothGrowth$resid <- rstudent(model_anova_2w)  
ToothGrowth$fitted <- fitted(model_anova_2w)

# Normality

ggplot(ToothGrowth, aes(sample = resid)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "Q-Q Plot of Residuals", 
       x = "Theoretical Quantiles", 
       y = "Sample Quantiles") +
  theme_minimal()

# Residuals vs Fitted plot to check variance patterns

ggplot(ToothGrowth, aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal() 


```

# Hands-on REPEATED MEASURES ANOVA

## WeightLoss: data 

Data on weight loss and self esteem over three months, for three groups of individuals: Control, Diet and Diet + Exercise.

```{r , warning=F,echo=TRUE, eval=TRUE}
library(tidyr)
library(carData)
library(dplyr)
library(gtsummary)
library(emmeans)

data("WeightLoss")

WeightLoss <- WeightLoss |>
    filter(group!="DietEx") |>  # Keep Control and Diet groups 
    mutate(group =factor(group),  # Group as factor  
    subject = factor(row_number())) |> # Create identifier as factor
    dplyr::select(-c(se1,se2,se3)) # Discard self esteem

WeightLoss_long <- WeightLoss |>
  pivot_longer(
    cols = starts_with("wl"),  # Select weight columns
    names_to = "Time",             # New column for time points
    values_to = "Weight"           # New column for weight values
  ) |>
  mutate(Time = factor(gsub("wl", "", Time)))  # Time as factor


```


## WeightLoss: data 


```{r , warning=F,echo=TRUE, eval=TRUE}

WeightLoss_long |> 
  dplyr::select(-subject) |>
  tbl_strata(
    strata = group,  .tbl_fun =  ~ .x |>
        tbl_summary(by = Time, 
        missing = "no",
        type = Weight ~ "continuous2", 
        statistic=all_continuous()~c("{mean} ({sd})",
        "{median} ({p25}, {p75})"),
        digits = all_continuous() ~ 2,
        label = list(Weight = "Diff in Kg")) )
            

```


## WeightLoss: data 


```{r , warning=F,echo=TRUE, eval=FALSE}


# Plot
ggplot(WeightLoss_long, aes(x = Time, y = Weight , color = group)) +
  geom_point(size = 4, alpha = 0.3) +
  geom_boxplot(aes(x = Time, group = interaction(Time, group)), width = 0.4, alpha = 0.5) +
  facet_wrap(~group) + 
  labs(x = "Time", y = "Diff in Kg") +
  theme_minimal() 

```


## WeightLoss: normality


```{r , warning=F,echo=TRUE, eval=FALSE}

#QQplot
ggplot(WeightLoss_long, aes(sample = Weight)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~Time*group) +  
  labs(title = "Q-Q Plots for Each Group",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()



```


## WeightLoss: anova


```{r , warning=F,echo=TRUE, eval=T}

# ANOVA: dependent ~ independent
model_anova_rm <- aov(Weight ~ Time + group + Error(subject/Time), data = WeightLoss_long)  

# Arrange results
results_anova_rm <- broom::tidy(model_anova_rm)
results_anova_rm$term<-c("Group","Residuals","Time","Residual")
names(results_anova_rm)<-c("Stratum","Term","df","SS","MSQ","F","P-value")


# Tabulate results
results_anova_rm |>
  gt::gt() |>
  gt::fmt_number(decimals = 3) |>
  gt::tab_header(title = "Weigth: repeated measure Anova") 

```


## WeightLoss: homocedasticity

**Mauchly's Test for Sphericity**


```{r , warning=F,echo=TRUE, eval=T}
library(ez)

anova_test <- ezANOVA(  
  data = WeightLoss_long,
  dv = Weight,      # Dependent variable
  wid = subject,    # Repeated measure (subject)
  within = Time,    # Within-subject factor
  between = group,  # Between-subject factor
  detailed = TRUE
)

anova_test[2:3]

```

GG: Greenhouse-Geisser epsilon; HFe: Huynh-Feldt epsilon.


## WeightLoss: marginal means


```{r , warning=F,echo=TRUE, eval=T}

emm_results <-emmeans(model_anova_rm, ~ Time | group)

pairs(emm_results, adjust = "tukey")

```


## WeightLoss: marginal means


```{r , warning=F,echo=TRUE, eval=T}

library(afex)

model_afex <- aov_car(Weight ~ Time + group + Error(subject/Time), data = WeightLoss_long)

emm_results <-emmeans(model_afex, ~ Time | group)

pairs(emm_results, adjust = "tukey")

```


## WeightLoss: marginal means

```{r , warning=F,echo=TRUE, eval=T}

library(afex) # Properly accounts for repeated measures

model_afex <- aov_car(Weight ~ Time + group + Error(subject/Time), data = WeightLoss_long)

emm_results <-emmeans(model_afex, ~ group | Time)

pairs(emm_results, adjust = "tukey")

```


## WeightLoss: marginal means


```{r , warning=F,echo=TRUE, eval=FALSE}


# Plot estimated marginal means with confidence intervals

as.data.frame(emm_results) |> 
ggplot(aes(x = Time, y = emmean, color = group, group = group)) +
  geom_point(size = 3) +
  geom_line() +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.2) +
  labs(x = "Time", y = "Estimated Mean Weight") +
  theme_minimal() 


```


## WeightLoss: anova + interaction


```{r , warning=F,echo=TRUE, eval=FALSE}

# ANOVA: dependent ~ independent
model_anova_rm <- aov(Weight ~ Time * group + Error(subject/Time), data = WeightLoss_long)  

# Arrange results
results_anova_rm <- broom::tidy(model_anova_rm)
results_anova_rm$term<-c("Group","Residuals","Time","Time:Group","Residual")
names(results_anova_rm)<-c("Stratum","Term","df","SS","MSQ","F","P-value")


# Tabulate results
results_anova_rm |>
  gt::gt() |>
  gt::fmt_number(decimals = 3) |>
  gt::tab_header(title = "Weigth: repeated measure Anova") 

```


## Final messages

- ANOVA is a powerful tool for analyzing experimental designs.

- The assumption of heterogeneity is key.

- Normality? Not so much...

- Post hoc analysis: handle with care.

- Surprise: ANOVA is a linear model!  




